VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-03 doc/publications/journal2014/paper.tex
comparison

Clustering the samples 30\% fastest in \variantSpark{} compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
The 30 second speed increase in \variantSpark{} is likely due us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors making operations less efficient.

We also compared the k-means clustering times to a multi-threaded approach in Python and R. 
Although an order of magnitude slower than \variantSpark{} and ADAM, k-means clustering in both Python and R were faster than our Hadoop implementation; Python required 11min and 29sec, whereas R required 7min and 25sec. 
This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine, which enables R and Python to store the output from each k-means iteration directly in memory whereas Hadoop is designed to write the output of each iteration to disk. 
This demonstrates the improvements of {\sc Spark} over Hadoop where datasets can be processed in memory, virtually eliminating the disk IO bottleneck on smaller datasets.

@@

\section*{Conclusions}

\section*{Materials and methods}
\subsection*{Computational resources}
We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB RAM and Ubuntu installed as the OS. 
For the whole-genome clustering, we used our in-house Hadoop cluster. This uses Hadoop 2.5.0, managed by CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.

\subsection*{VariantSpark Implementation}
For each of the implementations, we pre-process the VCF files to arrays of features.
In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD). RDDs allow us to process the files, line-by-line, in parallel. We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file.
The key-value pairs (KVPs) from each line in the VCF file are stored in separate arrays. We then `zip' each array with a unique index, where these indexes form identifiers for the alleles.
At this point, we have one array for each line from the VCF file. The values in the arrays are KVPs of the parsed tab-separated values zipped with their heading (individual ID), and each array has a unique index.
We now use a `flatMap' to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out KVPs that aren't alleles (i.e. KVPs that are VCF metadata).
This results in KVPs where the key is the individual ID, and the value is a tuple of the allele and it's index. We effectively have one KVP for each allele from the VCF file. At this point, each allele is still a string straight from the VCF file (i.e. `0\textbar1'), however, we require doubles.
To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant. Also, for each individual, we filter out any alleles that have no variant, leaving us with KVPs of variants (homozygous and heterozygous) only.
We can now group these KVPs by their key (the individual ID) and map them to sparse vectors. This results in a sparse vector for each individual, the format required by MLLibs machine learning algorithms.

@@

%% TODO %%
\subsection*{Personal Genome Project VCF making}
The PGP data.....

>>>>

Clustering the samples 30\% fastest in \variantSpark{} compared to ADAM (1min and 20secs compared to 1min and 52secs) and  %80 vs 112
90\% faster than Hadoop (14mins and 23secs), see Table~\ref{fivewaycomparison}.
The similarity between \variantSpark{} and ADAM is expected, as they both use {\sc Spark}'s MLlib k-means implementation. 
The 30 second speed increase in \variantSpark{} is likely due to us converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.

To demonstrate {\sc Spark}'s advantages over parallel approaches to k-means clustering in existing libraries, we compare the clustering libraries used in {\variantSpark}, to those in Python and R. 
In regards to speed, clustering using Python and R were both an order of magnitude slower than \variantSpark{} and ADAM, however both were faster than our Hadoop implementation, where Python required 11min and 29sec and R required 7min and 25sec. 
This is due the dataset (chromosome 22) being small enough to fit into memory on a single machine. This enables Python and R to store the output from each k-means iteration directly in memory. Hadoop, however, is designed to write the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue, the bottleneck introduced from disk IO increases processing time notably.
This demonstrates one of the major improvements of {\sc Spark} over Hadoop, where {\sc Spark} can cache datasets in memory. This virtually eliminates the disk IO bottleneck on smaller datasets, or at least reduces it on larger datasets where only a portion of the data can be cached in memory.

For smaller datasets, we've demonstrated {\sc Spark} to be faster than it's single-machine counterparts, however it's also important that this speedup isn't at the cost of quality.
The 1000 Genome data is ideal for investigating the clustering quality, as the consortium provides metadata which lists the super-population that each individual belongs to. This knowledge allows us to create a `truth' clustering, where individuals are clustered by their super-population.
With the four super-populations (AMR, AFR, EAS, SAS), we can divide the individuals up into four clusters based on these labels. We can then compare this truth clustering with the k-means clusters.
For this comparison, we use the Adjusted Rand Index (ARI) metric. To generate the ARI, an algorithm takes two separate clusterings and returns a value based on their similarity.
Two identical clusterings will result in a 1, whereas 0 represents two clusterings that do not agree on any points (individuals). On Chromosome 22, each or the algorithms resulted in an ARI of 0.84. Although not perfect, this demonstrates that the speed-up in spark isn't at the cost of quality, when compared to R and Python.
A result of 1 is probably unlikely because, as mentioned by 10.15252/embr.201439469, although differences in the distribution of alleles exist between populations, the boundaries are rather diffuse. %TODO - proper citation

@@

\section*{Conclusions}
By using the k-means to cluster the data, we can compare the results with

\section*{Materials and methods}
\subsection*{Computational resources}
We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory and Ubuntu as the OS. 
For the whole-genome clustering, we used our in-house Hadoop cluster. This uses Hadoop 2.5.0, managed by CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.

\subsection*{VariantSpark Implementation}
For each of the implementations, we pre-process the VCF files to arrays of features.
In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD). RDDs allow us to process the files, line-by-line, in parallel. We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file.
The key-value pairs (KVPs) from each line in the VCF file are stored in separate arrays. We then `zip' each array with a unique index, where these indexes form identifiers for the alleles.
At this point, we have one array for each line from the VCF file. The values in the arrays are KVPs of the parsed tab-separated values zipped with their heading (individual ID), and each array has a unique index.
We now use a `flatMap' to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out KVPs that aren't alleles (i.e. KVPs that are VCF metadata).
This results in KVPs where the key is the individual ID, and the value is a tuple of the allele and it's index. We effectively have one KVP for each allele from the VCF file. At this point, each allele is still a string straight from the VCF file (i.e. `0\textbar1'), however, we require doubles for MLLib.
To convert this to a double, we take the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant. Also, for each individual, we filter out any alleles that have no variant, leaving us with KVPs of variants (homozygous and heterozygous) only.
We can now group these KVPs by their key (the individual ID) and map them to sparse vectors. This results in a sparse vector for each individual, the format required by MLLibs machine learning algorithms.

@@

%% TODO (and better heading) %%
\subsection*{Personal Genome Project VCF making}
The PGP data.....
